---
title: "LAB6: Work with Models"
author: "Neskarly Marin"
format:
  revealjs:
    embed-resources: true
    theme: solarized
    slide-number: true
    chalkboard: false
    transition: fade
    code-fold: true
    code-line-numbers: false
    center: false
    width: 1200
    height: 675
    scrollable: true
    code-overflow: wrap
execute:
  echo: true
  warning: false
  message: false
knitr:
  opts_chunk:
    dev: "png"
    fig-align: "center"
---

```{r}
#| include: false
library(tidyverse)
library(ggplot2)
library(DataExplorer)
library(GGally)
library(broom)
library(margins)
library(gglm)
library(gridExtra)
```

```{r}
#| include: false
load("Catholic_viz.RData")

str(Catholic_viz)
summary(Catholic_viz)
```

# Remove ID Variable

```{r}
Catholic_viz <- Catholic_viz %>%
  select(-id, -asian) 
```

The ID variable was removed because it does not contain predictive information and only serves as an identifier.

# Convert Binary Variables to Factors

```{r}
Catholic_viz <- Catholic_viz %>%
mutate(
female = as.factor(female),
hispan = as.factor(hispan),
black = as.factor(black),
hsgrad = as.factor(hsgrad),
cathhs = as.factor(cathhs),
parcath = as.factor(parcath)
)
```

Binary variables were converted to factors to improve interpretability in visualizations and regression analysis.

# Exploratory Data Analysis (EDA)

## Dataset Overview

```{r}
plot_intro(Catholic_viz)
```

### Insights

::: {style="font-size: 0.65em;"}
The dataset contains 5,970 observations and multiple variables describing student academic performance, demographics, parental education, and family income.
:::

## Missing Values

```{r}
plot_missing(Catholic_viz)
```

### Insights

::: {style="font-size: 0.65em;"}
The missing value plot indicates that there are no missing values, suggesting that the dataset is suitable for regression analysis.
:::

## Nominal Variables

::::: columns
::: {.column width="75%"}
```{r}
plot_bar(Catholic_viz)
```
:::

::: {.column width="25%" style="font-size: 0.65em; line-height: 1.4;"}
### Insights

-   Gender: Fairly balanced, with slightly more females.

-   Demographics: The sample is predominantly non-Hispanic and non-Black.

-   Education & Religion: Almost all are high school graduates. Few attended Catholic high schools, but a large portion have at least one Catholic parent.
:::
:::::

## Distribution of Numeric Variables

::::: columns
::: {.column width="75%"}
```{r}
#| fig-width: 12
#| fig-height: 8
plot_histogram(Catholic_viz)
```
:::

::: {.column width="25%" style="font-size: 0.65em; line-height: 1.4;"}
### Insights

-   **Test Scores:** Both are slightly left-skewed, with most scores peaking in the upper 50s.

-   **Parental Education:** Shows discrete spikes at 12, 14, and 16 years, clearly aligning with major educational milestones (high school, college).

-   **Family Income:** Normally distributed around 10.5. The values suggest it is already log-transformed, which is ideal for linear regression.
:::
:::::

## Boxplots for numeric variables

::::: columns
::: {.column width="75%"}
```{r}
#| fig-width: 9
#| fig-height: 4.5
Catholic_viz$All <- "All"
plot_boxplot(Catholic_viz, by = "All")
Catholic_viz <- Catholic_viz %>% select(-All)
```
:::

::: {.column width="25%" style="font-size: 0.65em; line-height: 1.4;"}
### Insights

-   **Test Scores:** Symmetric distributions with centered medians and no extreme outliers, which is ideal for regression modeling.

-   **Parental Education:** Highly concentrated between 12 and 14 years, displaying a few distinct outliers at the extreme ends.

-   **Family Income:** Tightly clustered around the median, though the plot clearly reveals a left tail of low-income outliers that were not as obvious in the histogram.
:::
:::::

## Correlations

::::: columns
::: {.column width="75%"}
```{r}
#| fig-width: 9
#| fig-height: 4.5
plot_correlation(Catholic_viz, type="continuous")
```
:::

::: {.column width="25%" style="font-size: 0.65em; line-height: 1.4;"}
### Insights

-   Math scores have a strong positive correlation (0.71) with reading scores.

-   Father's education (0.32), family income (0.27), and mother's education (0.27) all have weak-to-moderate positive relationships with reading scores.

-   **Multicollinearity Check:** Mother's and father's education are moderately correlated with each other (0.58). This is expected and is not high enough to cause major multicollinearity issues in your regression model.
:::
:::::

## Scatterplot

::::: columns
::: {.column width="75%"}
```{r}
#| fig-width: 9
#| fig-height: 4.5
ggplot(Catholic_viz, aes(x = math12, y = read12)) +
geom_point(alpha = 0.15) +
geom_smooth(method = "lm", aes(color = "OLS")) +
geom_smooth(method = "loess", aes(color = "LOESS")) +
scale_color_manual(name = "Model", values = c("OLS"="red","LOESS"="blue")) +
theme_minimal() +
labs(
title="Relationship Between Math and Reading Scores",
x="Math Score (12th Grade)",
y="Reading Score (12th Grade)"
)
```
:::

::: {.column width="25%" style="font-size: 0.65em; line-height: 1.4;"}
### Insights

-   There is a strong, positive relationship between math and reading scores. As math scores increase, reading scores reliably increase as well.

-   The LOESS curve (blue) almost perfectly overlaps the OLS line (red). This confirms that the relationship is strictly linear.

-   The points are fairly evenly distributed around the regression lines
:::
:::::

# Modeling

## Multiple Linear Regression

```{r}
model <- lm(read12 ~ math12 + fatheduc + lfaminc, data = Catholic_viz)

summary(model)
```

### Insights

::: {style="font-size: 0.65em;"}
-   The final model explains 51% of the variance in reading scores.

-   Math score is by far the strongest predictor. Holding everything else constant, for every 1-point increase in a student's math score, their reading score goes up by about 0.68 points.

-   Father's education and family income are also significant predictors.
:::

## Predicted vs Actual

```{r}
plot_data <- augment(model)

ggplot(plot_data, aes(x = .fitted, y = read12)) +
geom_point(alpha = 0.3) +
geom_abline(slope = 1, intercept = 0, color = "red") +
theme_minimal() +
labs(
title="Predicted vs Actual Reading Scores",
x="Predicted Reading Scores",
y="Actual Reading Scores"
)
```

### Insights

::: {style="font-size: 0.65em;"}
-   Points cluster consistently along the red reference line, confirming the model captures the overall trend accurately.
:::

## Residual Diagnostics

::::: columns
::: {.column width="75%"}
```{r}
#| fig-width: 9
#| fig-height: 4.5
gglm(model)
```
:::

::: {.column width="25%" style="font-size: 0.65em; line-height: 1.4;"}
### Insights

-   The Normal Q-Q plot shows the data points closely following the diagonal line, confirming that the model's residuals are approximately normally distributed.

-   The Residuals vs. Fitted and Scale-Location plots display a relatively random scatter of points with mostly flat trend lines.

-   The Residual vs. Leverage plot reveals no extreme outliers with high leverage. This means no single data point is disproportionately skewing our model's results.
:::
:::::

## Variable importance

::::: columns
::: {.column width="75%"}
```{r}
#| fig-width: 9
#| fig-height: 4.5
tidy(model) %>%
filter(term != "(Intercept)") %>%
ggplot(aes(x=reorder(term, estimate), y=estimate)) +
geom_col() +
coord_flip() +
theme_minimal() +
labs(
title="Regression Coefficients for Reading Score Model",
x="Predictor",
y="Coefficient Estimate"
)
```
:::

::: {.column width="25%" style="font-size: 0.65em; line-height: 1.4;"}
### Insights

-   The coefficient for math scores is by far the largest, visually confirming it as the most dominant predictor of reading scores in the analysis.

-   Family income and father's education display moderate positive coefficients, highlighting their secondary role in predicting reading performance.

-   Mother's education and Catholic high school attendance show the smallest coefficient estimates, aligning with their lack of statistical significance.
:::
:::::

## Marginal Effects

::::: columns
::: {.column width="75%"}
```{r}
#| fig-width: 9
#| fig-height: 4.5
marg <- margins(model)

summary(marg)

plot(marg)
```
:::

::: {.column width="25%" style="font-size: 0.65em; line-height: 1.4;"}
### Insights

-   Because this is a standard linear model, the Average Marginal Effects (AME) perfectly mirror our regression coefficients. They confirm the expected, uniform change in reading scores for every one-unit increase in each predictor.
:::
:::::

# Final Reflection

::: {style="font-size: 0.75em;"}
-   Once again, DataExplorer made the initial data checks incredibly efficient. Instead of guessing, I could immediately see the distributions and correlations, which guided my variable selection from the very beginning.
-   A good model isn't about throwing every available variable at the wall to see what sticks. Reviewing the p-values and dropping the non-significant predictors taught me the value of building a clean model.
-   Generating the Predicted vs. Actuals and gglm plots helped me actually see what an R-squared of 0.51 looks like in reality.
:::
